---
title       : An Analysis of Motorcycle Safety
subtitle    : Why my girlfriend should let me ride a motorcycle
author      : Dan Garant
job         : 
framework   : io2012        # {io2012, html5slides, shower, dzslides, ...}
highlighter : highlight.js  # {highlight.js, prettify, highlight}
hitheme     : tomorrow      # 
widgets     : []            # {mathjax, quiz, bootstrap}
mode        : selfcontained # {standalone, draft}
---

```{r, echo=FALSE, cache=TRUE, warning=FALSE}
r <- read.csv("C:/Users/Dan/Downloads/MData.tsv", sep="\t", 
              stringsAsFactors = FALSE, colClasses="numeric", header=TRUE)
r$motorcycle <- ifelse(r$restraint %in% c(5, 16, 17), "Motorcycle", "Not Motorcycle")
r$restrained <- with(r, (restraint %in% c(0, 1, 2, 3, 4, 5, 8, 10, 11, 12, 97, 98, 99)))  # restraints / helmets used
r$not.speeding <- with(r, speeding == 0) # not speeding
r$no.drugs <- with(r, (alcres < 1 | alcres == 96) & druginv %in% c(-1, 0, 8, 9) & malcohol %in% c(0, 8, 9))
r$complied <- with(r, licrestrict != 2 & prevdwi %in% c(0, 99))
r$not.impaired <- with(r, dricondatcrashtime %in% c(0, 98, 99) & dridistract %in% c(0, 96, 99))
r$responsible <- with(r, not.speeding & no.drugs &  restrained &  complied & not.impaired)
r$conditions <- with(r,
                  ifelse(
                  atmcond %in% c(0, 1, 8, 10, 98, 99) & # no adverse atmospheric conditions
                  dayofweek >= 2 & dayofweek <= 6 & # during the week
                  acctime >= 500 & acctime <= 1900 & # during daylight hours      
                  rfun < 10 # roadway is rural
                  , "Good Motorcycling Conditions", 
                  "Urban / Poor Weather / Dark"))

library(ggplot2)
library(plyr)
library(ggthemes)

count.by.motor <- table(r$motorcycle)
reg <- c(257512000, 6567197)
names(reg) <- c("Not Motorcycle", "Motorcycle")
```

## Legend vs. Reality

- Many people believe motorcycles are much more dangeous than cars
- Undoubtedly cyclists are more exposed and harder to see
- However, cyclists are also more alert
- It's clear from traffic statistics that cyclists are disproportionately involved in fatal accidents
  - http://www-fars.nhtsa.dot.gov/
- Before we jump to conclusions about motorcycle safety, let's understand why this is


--- .class #id 

## The Data

- The National Highway Traffic Safety administration publishes a yearly data set describing fatal accidents occurring within that year
- For each accident, a number of factors are reported including weather conditions, time of day, whether safety devices were used, whether speeding or drugs were involved, etc.
- This presentation uses the 2011 data set, consisting of 67,143 accident reports



---

## A Note About Relative Frequency

- Without a doubt, there are more fatalities in accidents involving cars, trucks, and SUVs than there are involving motorcycles, so motorcycle accidents are less __frequent__
- However, there are fewer registered motorcycles on the road, so the __relative frequency__ of motor accidents is actually much higher


```{r, echo=FALSE, fig.width=14, fig.height=4, cache=TRUE, fig.align='center'}
theme_set(theme_solarized(base_size=16))
library(gridExtra)
library(plyr)
library(ggplot2)
g1 <- ggplot(r, aes(x=motorcycle,)) + geom_bar() + labs(y="Number of Fatalities", x="Vehicle Type", title="Frequency of Fatalities by Vehicle Type")
d <- ddply(r, .(motorcycle), summarize, rel.freq = length(motorcycle) / reg[motorcycle[1]])
g2 <- ggplot(d, aes(x=motorcycle, y=rel.freq)) + geom_bar(stat="identity") + labs(y="Relative Frequency of Fatality", x="Vehicle Type", title="Relative Frequency of Fatalities by Vehicle Type")
grid.arrange(g1, g2, ncol=2)
```

- So, let's get started ...

---

## Helmet / Seatbelt Use

Over 70% of 4-wheeled vehicle operators use seat belts, but less than 40% of motorcyclists wear helmets

```{r echo=FALSE, fig.align='center', fig.width=7, fig.height=6, cache=TRUE}
theme_set(theme_solarized(base_size=16))
d <- ddply(r, .(motorcycle), summarize,
           pct.restrained=(length(restrained[restrained]) /  length(motorcycle)))
g <- ggplot(d, aes(x=motorcycle, y=pct.restrained)) + 
      geom_bar(stat="identity") +
    labs(y="Percentage Using Restraints / Helmets", title="Restraints / DOT-Approved Helmets Used", x="Vehicle Type")
print(g)
```


---

## Drug and Alcohol Usage

Motorcycle accidents are almost 15% more likely to involve drugs or alcohol than other accidents

```{r echo=FALSE, fig.align='center', fig.width=7, fig.height=6, cache=TRUE}
theme_set(theme_solarized(base_size=16))
d <- ddply(r, .(motorcycle), summarize,
           pct.drug.free=(length(no.drugs[no.drugs]) / length(motorcycle)))
g <- ggplot(d, aes(x=motorcycle, y=(1-pct.drug.free))) + 
      geom_bar(stat="identity") + 
      labs(y="Percentage Under Influence of Drugs/Alc.", x="Vehicle Type",
           x="Drug and Alcohol Usage")
print(g)
```


---

## Speeding

Motorcycle accidents are about 10% more likely to have speeding as a factor, as indicated by police reports

```{r echo=FALSE, fig.align='center', fig.width=7, fig.height=6, cache=TRUE}
theme_set(theme_solarized(base_size=16))
d <- ddply(r, .(motorcycle), summarize,
           pct.not.speeding=(length(not.speeding[not.speeding]) / length(motorcycle)))
g <- ggplot(d, aes(x=motorcycle, y=1-pct.not.speeding)) + 
  geom_bar(stat="identity") + 
  labs(y="Percentage Speeding", x="Vehicle Type", title="Speeding Fatalities")
print(g)
```


---

## Legal Motorists

Many more motorcyclists than other operators are not legally permitted to drive their vehicles, 
either because they never got a license, or it was revoked or suspended.

```{r echo=FALSE, fig.align='center', fig.width=7, fig.height=6, cache=TRUE}
theme_set(theme_solarized(base_size=16))
d <- ddply(r, .(motorcycle), summarize,
           pct.complied=(length(complied[complied]) / length(motorcycle)))
g <- ggplot(d, aes(x=motorcycle, y=1-pct.complied)) +
  geom_bar(stat="identity") + 
  labs(y="Percentage With Suspended / Revoked Licenses", x="Vehicle Type", title="Fatalities by License Compliance")
print(g)
```

---

## Weather Conditions

Most motorcycle accidents occur in non-ideal motorcycle conditions, such as 
urban environments, in poor weather, or in the dark.

```{r echo=FALSE, fig.align='center', fig.width=7, fig.height=6, cache=TRUE}
theme_set(theme_solarized(base_size=16))
d <- ddply(subset(r, motorcycle == "Motorcycle"), .(conditions), summarize,
           num.accidents = length(conditions))
g <- ggplot(d, aes(x=conditions, y=num.accidents)) + 
  geom_bar(stat="identity") +
  labs(x="Weather Conditions", y="Number of Accidents", 
       title="Fatalities by Driving Conditions")
print(g)
```

---

## So What?

All of these factors are under the control of the cyclist.
Let's see how things break down for a responsible cyclist 
that only rides under ideal conditions.

Responsibility Factors:
 - Always wear a helmet
 - Don't use drugs/alcohol
 - Don't speed
 - Don't ride with a revoked/suspended license

Condition Factors:
 - Don't ride in congested areas (e.g. cities)
 - Don't ride in bad weather
 - Don't ride in the dark


--- 

## Judge the Risk for Yourself

```{r echo=FALSE, fig.align='center', fig.width=13, fig.height=7, cache=TRUE}
theme_set(theme_solarized(base_size=16))
r$responsiblewords <- ifelse(r$responsible, "Responsible", "Irresponsible")
d <- ddply(subset(r), 
           .(responsiblewords, motorcycle, conditions), summarize, 
           fatalities.per.10000=(length(motorcycle) / reg[motorcycle[1]]) * 10000)
g <- ggplot(d, aes(x=motorcycle, y=fatalities.per.10000)) +
        facet_grid(responsiblewords~conditions) + geom_bar(stat="identity") + labs(x="Vehicle Type", 
              y="Fatalities per 10,000 Registered Vehicles",
              title="Fatalities by Driver Responsibility and Driving Condition")
print(g)
```






